---
title: "Week 3: the `tidyverse`"
subtitle: |
  | GR5072 - Modern Data Structures
  | Spring 2022 
output:
  md_document:
      variant: markdown_github
---

</br>

### Instructions 

This is a notebook with starter code for you to follow the examples that we'll review in class. This is a place for you to add your class notes. *REMEMBER:* `push`ing back this annotated-by-you notebook and the associated GitHub-flavored markdown file to your **GitHub Classroom** repo counts towards you class participation grade.

</br>

### Loading packages

Let's begin by loading all the packages that we'll need today

[INSERT YOUR NOTES HERE]

```{r packages}
library(tidyverse)
library(nycflights13)
```


### Creating Tibbles

[INSERT YOUR NOTES HERE]

```{r}

tibble(x = letters)
```


## Printing

[INSERT YOUR NOTES HERE]

```{r}
tibble(x = -5:1000)
```

## Subsetting

[INSERT YOUR NOTES HERE]

```{r}
df1 <- data.frame(x = 1:3, y = 3:1)
class(df1[, 1:2])

class(df1[, 1])
```

[INSERT YOUR NOTES HERE]

```{r}
df2 <- tibble(x = 1:3, y = 3:1)
class(df2[, 1:2])

class(df2[, 1])
```

## To extract a single column use `[[` or `$`:

[INSERT YOUR NOTES HERE]

```{r}
class(df2[[1]])

class(df2$x)
```

## Intro to `dplyr` package 

[INSERT YOUR NOTES HERE]

```{r}
## using data contained in the nycflights13 package that we loaded at the beginning of this notebook
dim(flights)

flights
```

## Single table verbs

[INSERT YOUR NOTES HERE]

## Filter rows with `filter()`

[INSERT YOUR NOTES HERE]

```{r}
filter(flights, month == 1, day == 1)
```

## Filter rows with `filter()`

[INSERT YOUR NOTES HERE]

```{r}
flights[flights$month == 1 & flights$day == 1, ]
```


## Arrange rows with `arrange()`

[INSERT YOUR NOTES HERE]

```{r}
arrange(flights, year, month, day)
```

[INSERT YOUR NOTES HERE]

```{r}
arrange(flights, desc(arr_delay))
```

## Select columns with `select()`

[INSERT YOUR NOTES HERE]

## Select columns by name

[INSERT YOUR NOTES HERE]

```{r}
select(flights, year, month, day)
```

## Select all columns between year and day (inclusive)

[INSERT YOUR NOTES HERE]

```{r}
select(flights, year:day)
```

## Select all columns except those from year to day (inclusive)

[INSERT YOUR NOTES HERE]

```{r}
select(flights, -(year:day))
```

## Helper functions

[INSERT YOUR NOTES HERE]

## Use `rename()`  to change variable names

[INSERT YOUR NOTES HERE]

```{r}
rename(flights, tail_num = tailnum)
colnames(flights)
```

## Add new columns with `mutate()`

[INSERT YOUR NOTES HERE]

```{r}
mutate(flights,
  gain = arr_delay - dep_delay,
  speed = distance / air_time * 60
)
```


[INSERT YOUR NOTES HERE]


```{r}
starwars %>%
  select(name:mass, gender, species) %>%
  mutate(
    type = case_when(
      height > 200 | mass > 200 ~ "large",
      species == "Droid"        ~ "robot",
      TRUE                      ~ "other"  
      # this "TRUE" code defines what to define everything                                              # else
    )
  )
```

## Summarise values with `summarise()`


[INSERT YOUR NOTES HERE]


```{r}
summarise(flights,
  delay = mean(dep_delay, na.rm = TRUE)
)
```

## Calculate stats by variable category

[INSERT YOUR NOTES HERE]

## `group_by()` examples:

[INSERT YOUR NOTES HERE]

```{r}
destinations <- group_by(flights, dest)
summarise(destinations,
  planes = n_distinct(tailnum),
  flights = n()
)
```


## Pipe Operators and the `tidyverse`

[INSERT YOUR NOTES HERE]

## Different ways to organize code

[INSERT YOUR NOTES HERE]

```{r}
a <- filter(mtcars, carb > 1)
b <- group_by(a, cyl)
c <- summarise(b, Avg_mpg = mean(mpg))
d <- arrange(c, desc(Avg_mpg))
print(d)
```


[INSERT YOUR NOTES HERE]

```{r}
arrange(
   summarize(
       group_by(
           filter(mtcars, carb > 1),
           cyl
          ),
       Avg_mpg = mean(mpg)
      ),
   desc(Avg_mpg)
 )
```


[INSERT YOUR NOTES HERE]

```{r}
mtcars %>%
        filter(carb > 1) %>%
        group_by(cyl) %>%
        summarise(Avg_mpg = mean(mpg)) %>%
        arrange(desc(Avg_mpg))
```



## Separating

[INSERT YOUR NOTES HERE]

```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")
```

[INSERT YOUR NOTES HERE]

```{r}
#Separate year values after second element of value

table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```

## Uniting

[INSERT YOUR NOTES HERE]

```{r}
table5[1:2,]

table5 %>% 
  unite(new, century, year, sep = "")
```

## Joining (merging) data in the Tidyverse

[INSERT YOUR NOTES HERE]

## Keys

[INSERT YOUR NOTES HERE]


```{r}
planes %>% 
  count(tailnum) %>% 
  filter(n > 1)
```


## Mutating Joins

[INSERT YOUR NOTES HERE] 

```{r}
flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)
flights2
```

[INSERT YOUR NOTES HERE]

```{r}
# the airlines dataset looks like:
head(airlines)
```

[INSERT YOUR NOTES HERE]

```{r}
flights2 %>%
  select(-origin, -dest) %>% 
  left_join(airlines, by = "carrier")
```

[INSERT YOUR NOTES HERE]

```{r}
flights2 %>%
  select(-origin, -dest) %>% 
  left_join(airlines, by = "carrier")
```

## Inner Join

[INSERT YOUR NOTES HERE]

```{r}
x <- tribble(
  ~key, ~val_x,
     1, "x1",
     2, "x2",
     3, "x3"
)
y <- tribble(
  ~key, ~val_y,
     1, "y1",
     2, "y2",
     4, "y3"
)
```

[INSERT YOUR NOTES HERE]

```{r}
x %>% 
  inner_join(y, by = "key")
```

## Outer joins

[INSERT YOUR NOTES HERE]

```{r}
#leftjoin() with same variable name in two datasets
flights2 %>% 
  left_join(planes, by = "tailnum")
```

[INSERT YOUR NOTES HERE]

```{r}
#leftjoin() with diff't variable name in two datasets
flights2 %>% 
  left_join(airports, c("dest" = "faa"))
```


## Gathering

[INSERT YOUR NOTES HERE]

```{r}
table4a
```

[INSERT YOUR NOTES HERE]


```{r}
# Note that columns are numeric, so we use backticks
# More typically we can just refer to column names.
table4a %>% 
  gather(`1999`, `2000`, key = "year", value = "cases")
```

## Spreading

[INSERT YOUR NOTES HERE]

```{r}
table2 %>%
    spread(key = type, value = count)
```
